\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{SmarteRun: Coding The Portable Coach}

### 2014-05-12 final/article.tex
Add results and analysis

@@

\usepackage{fancyhdr} % Headers and footers
\pagestyle{fancy} % All pages have headers and footers
\fancyhead{} % Blank out the default header
\fancyfoot{} % Blank out the default footer
\fancyhead[C]{Running title $\bullet$ November 2012 $\bullet$ Vol. XXI, No. 1} % Custom header text
\fancyfoot[RO,LE]{\thepage} % Custom footer text

@@

\lettrine[nindent=0em,lines=3]{B} eyond food, drink, and sleep, the next most essential facet of life is physical exercise. The most common form of exercise is running, but many find it hard to truly stick to a running regime or properly challenge themselves. People who lift weights often stick to rules with concrete results such as increasing their maximum lift every 2 weeks by 5\%, but running is much more difficult to systematize with quantifiable deliverables as it isn't broken up into repetitions or sets. Other than sprinters, a run is a run. A person leaves the start of the run, then at some point later, they end their run. Without someone watching the runner and giving feedback as to how they are doing, little progress is likely to come out of this exercise. \\
This is a problem many are interested in solving. With the insurgence of mobile apps, a number of them have reached success in providing real-time and even overall summarizing data during and at the end of a run. RunKeeper and MapMyRun are the two top apps, but they don't provide much more than a map of where the user has run with an instantaneous and average speed. There are no apps out there which give normative data on how a runner should be running to reach goals. In this day and age with machine learning capabilities surpassing many humans, there should be an app which watches how the user runs and, using data from past runs and the current run's parameters, recommends how a runner should proceed from that moment. Additionally, it would be good if the app provide motivation or somehow otherwise encouraged the runner to maintain the recommended pace. \\
	SmarteRun is a project attempting to address these needs. By using GPS data of a person's run, the app keeps track of each run's speed vs. time relation as well as the total length of the run. Then, during the run, the app uses machine learning to combine data from past recorded runs and determine a reasonable speed the runner should hold for the rest of the run. This project is about the investigation into how to properly combine past data and effectively deliver a proper recommendation speed. The intent is that this speed would be translated into a pace and ultimately used to decide a song that should be played to subliminally push the runner, thereby giving them the coach they don't even realize they have.

@@

\section{Background}
\indent The original idea for this project came from an observation that most runners listen to music. The reason most do is probably because music is entertaining and something to think about while running. It turns out that this distraction is more than just entertainment for an unfocused runner. In an article from the Journal of Sport Behavior, they discuss that music can actually allow for improved performance. There are a few basic reasons for this enhanced performance. \\

@@

	Given this research, we have validated the idea of using past data to predict future run performance as well as using music to improve and influence the runner'Äôs experience.

@@

	Using this predicted graph R(x,y,z) we provide tangent-line and secant-line predictions such that we provide a recommended pace for a runner such that given their current performance they achieve, exceed, or relax compared to their usual performance.  This step essentially merges our predictive data with the runner'Äôs current activity and would be the useful user-facing data provided in a retail application of our algorithms.

\subsection{Design Decisions}
	We consider again some choice design decisions made in the creation of our algorithms.  We note that for the scope of this assignment there were both many logical and arbitrary invariants we had to consider which influenced our design decisions.\\
	We determined that distance elapsed is the best metric of run progress after considering the alternative instantaneous speed.  It is difficult to define the term instantaneous as to produce speed a rate must be taken and therefore it is imperative to decide upon a time interval which we define as instantaneous.  Using elapsed distance allows the bypassing of this arbitrary design problem, though speed could still easily be calculated by measuring the change in elapsed distance over some defined time interval.\\
	We consider the tradeoff of with how many variables we represent a run and how tractable and verifiable our analysis is.  We considered a representation of n >= 4 dimensions, adding instantaneous parameters such as pace, heart rate, etc. and found that these additional parameters gave us little benefit while greatly complicating our analysis.  We found that variables time, distance elapsed, and total distance provided the simplest reasonable and most tractable effective results.\\
	To implement a k-nearest neighbors algorithm we had to make a best definition of nearest and determine how many neighbors to include in our analysis.  We considered other definitions of nearest but settled on comparing z-values as we assumed the invariant of good data and only comparing among the same runner such that many other variables can be ignored for the sake of our analysis.  We semi-arbitrarily came to the conclusion of using k=5 for our neighbors, determining that 5 strikes a nice balance between an accurate prediction, fast execution, and readable results.  We found that for k>=5 the quantitative benefit of additional neighbors dropped off exponentially.\\
	We decided to use secant and tangent lines in supplying our pace recommendation output due to their simplicity and understandability. For the purposes of our algorithms, we wanted to provide a simple metric which a runner could use as an immediate recommendation.  We considered outputting our more detailed itineraries though these are beyond the goals of our project.

@@

@@

\indent This project used a K-nearest-neighbor search attempting to take past runs and combine them in reveal information about how a typical run from a user would go. This estimation is then useful in predicting a good recommendation for how the runner should progress from that moment in the run. This approach to this issue resulted in a good prediction engine for short runs and a good starting point for longer runs. The longer runs are exceedingly difficult to extrapolate to because of the many factors that can come into play over the remaining time limit as well as the limited amount of long distance run data available during this project. Having access to a good amount of RunKeeper's or MapMyRun's data would allow for much more analysis. Combining run results is a difficult task and has many points of improvement that could probably apply to non-running type exercises as well. \\
	This investigation provided useful results, but it was intended mainly as a starting point. The problem of workout recommendation is the source of entire professions (i.e. personal trainers). This project assumed all runs were 'Äúgood'Äù runs and thus should be taken into account for prediction. In reality, some runs are ruined by a bad cramp, an error in data collection, or a phone call mid-run and thus should not be used to describe typical performance. Additionally, we ran our K-nearest-neighbor algorithm on the current run and matched other runs together based on just the Z variable of total run length. This could be improved if we tried to find runs similar to the current one based on the total length as well as the performance so far. This would allow us to, at least to some degree, more properly predict the longer runs as well as fix another major development that this problem needs before anyone can call it solved: atrophy during extended periods of inactivity. This app currently assumes the user is constantly training, however, a very common use case is ignored. Namely, a runner will often take a few weeks or even months off because they are particularly busy or it'Äôs winter. This would lead to a mismatch where the runner is in much worse shape than our current algorithm understands and then tells the runner to essentially sprint the rest of the run. Matching by performance so-far in the run would allow the algorithm to implicitly match on conditioning level as well. With a large enough user base, this could become very powerful as thousands of runs can be combined from a multitude of different users, much in the same way Netflix powers their suggestion engine. Furthermore, these approaches could easily be applied to many more types of exercise such as swimming or weight lifting as well. \\
	With a little bit of improvement, this type of project would be very valuable to an application such as RunKeeper. Taking the expected performance and translating it into a pace that the runner should keep would be a good next step in this project. Using assumptions based on the height of the runner would be a simple way to suggest a running pace based on speed, however an entirely separate study should be done on how to use machine learning to match a specific user to the proper pace based on where they are in the run and any other factors specific to their person and their run. This could then be a powerful tool to give runners the coach they don'Äôt even know they have.

>>>>

\usepackage{graphicx}

@@

\usepackage{fancyhdr} % Headers and footers
\pagestyle{fancy} % All pages have headers and footers
\fancyhead{} % Blank out the default header
\fancyfoot{} % Blank out the default footer
\fancyhead[C]{CS144 Spring 2014: Final Project, SmarteRun} % Custom header text
\fancyfoot[RO,LE]{\thepage} % Custom footer text

@@

\lettrine[nindent=0em,lines=3]{B} eyond food, drink, and sleep, the next most essential facet of life is physical exercise. The most common form of exercise is running, but many find it hard to truly stick to a running regime or properly challenge themselves. People who lift weights often stick to rules with concrete results such as increasing their maximum lift every 2 weeks by 5\%, but running is much more difficult to systematize with quantifiable deliverables as it isn't broken up into repetitions or sets. Other than sprinters, a run is a run. A person leaves the start of the run, then at some point later, they end their run. Without someone watching the runner and giving feedback as to how they are doing, little progress is likely to come out of this exercise. \\

This is a problem many are interested in solving. With the insurgence of mobile apps, a number of them have reached success in providing real-time and even overall summarizing data during and at the end of a run. RunKeeper and MapMyRun are the two top apps, but they don't provide much more than a map of where the user has run with an instantaneous and average speed. There are no apps out there which give normative data on how a runner should be running to reach goals. In this day and age with machine learning capabilities surpassing many humans, there should be an app which watches how the user runs and, using data from past runs and the current run's parameters, recommends how a runner should proceed from that moment. Additionally, it would be good if the app provide motivation or somehow otherwise encouraged the runner to maintain the recommended pace. \\

	SmarteRun is a project attempting to address these needs. By using GPS data of a person's run, the app keeps track of each run's speed vs. time relation as well as the total length of the run. Then, during the run, the app uses machine learning to combine data from past recorded runs and determine a reasonable speed the runner should hold for the rest of the run. This project is about the investigation into how to properly combine past data and effectively deliver a proper recommendation speed. The intent is that this speed would be translated into a pace and ultimately used to decide a song that should be played to subliminally push the runner, thereby giving them the coach they don't even realize they have.

@@

\section{Background}
\indent The original idea for this project came from an observation that most runners listen to music. The reason most do is probably because music is entertaining and something to think about while running. It turns out that this distraction is more than just entertainment for an unfocused runner. In an article from the Journal of Sport Behavior, they discuss that music can actually allow for improved performance. There are a few basic reasons for this enhanced performance. \\

@@

	Given this research, we have validated the idea of using past data to predict future run performance as well as using music to improve and influence the runner'Äôs experience. \\

@@

	Using this predicted graph R(x,y,z) we provide tangent-line and secant-line predictions such that we provide a recommended pace for a runner such that given their current performance they achieve, exceed, or relax compared to their usual performance.  This step essentially merges our predictive data with the runner'Äôs current activity and would be the useful user-facing data provided in a retail application of our algorithms.\\

\subsection{Design Decisions}
	We consider again some choice design decisions made in the creation of our algorithms.  We note that for the scope of this assignment there were both many logical and arbitrary invariants we had to consider which influenced our design decisions.\\

	We determined that distance elapsed is the best metric of run progress after considering the alternative instantaneous speed.  It is difficult to define the term instantaneous as to produce speed a rate must be taken and therefore it is imperative to decide upon a time interval which we define as instantaneous.  Using elapsed distance allows the bypassing of this arbitrary design problem, though speed could still easily be calculated by measuring the change in elapsed distance over some defined time interval.\\

	We consider the tradeoff of with how many variables we represent a run and how tractable and verifiable our analysis is.  We considered a representation of n >= 4 dimensions, adding instantaneous parameters such as pace, heart rate, etc. and found that these additional parameters gave us little benefit while greatly complicating our analysis.  We found that variables time, distance elapsed, and total distance provided the simplest reasonable and most tractable effective results.\\

	To implement a k-nearest neighbors algorithm we had to make a best definition of nearest and determine how many neighbors to include in our analysis.  We considered other definitions of nearest but settled on comparing z-values as we assumed the invariant of good data and only comparing among the same runner such that many other variables can be ignored for the sake of our analysis.  We semi-arbitrarily came to the conclusion of using k=5 for our neighbors, determining that 5 strikes a nice balance between an accurate prediction, fast execution, and readable results.  We found that for k>=5 the quantitative benefit of additional neighbors dropped off exponentially.\\

	We decided to use secant and tangent lines in supplying our pace recommendation output due to their simplicity and understandability. For the purposes of our algorithms, we wanted to provide a simple metric which a runner could use as an immediate recommendation.  We considered outputting our more detailed itineraries though these are beyond the goals of our project.\\

@@

\indent Given that the goal of the algorithm is predicting future pace information given current time and distance, we chose to assess the results quantitatively. The analysis involved taking a sample run (not in our database), calculating expected pace at a random point, and comparing the accuracy of our expected pace to the actual pace.\\

Here we present the full approach for measuring accuracy:
\begin{itemize}
\item Select a random run not used in the training data set
\item Select a random time point in that run, and determine the (time, distance) tuple for that point in time
\item Calculate the secant line between this time point and the end of the run, the slope of which represents average pace for the rest of the run
\item Now use the SmarteRun algorithm to model a predicted run using this same time, distance, and total distance
\item Calculate the secant line between this time point and the end of the run in this predicted model, the slope of which represents predicted average pace for the rest of the run

Repeat this process over multiple iterations, recording both the predicted pace and actual pace for each sample. Then calculate some measure of error (such as RMSE)
\end{itemize}

The purpose of this approach was to compare the accuracy of predicted pace to the actual pace for a run. Since our algorithm yields a full model of distance over time for a run, it is possible to sample multiple points along this model and compare expected pace to actual pace. We chose to use RMSE to measure accuracy because this project represents baseline accuracy; as we make changes to the algorithm in the future, we can optimize parameters with respect to minimizing RMSE.\\

Overall the results indicated a RMSE of 2.3088. As can be seen in the figure below, in general average pace was close to predicted pace (as shown by the large cluster of data points along the line $x=y$). However, clear groups of outliers exist, particularly in predicting too fast of a pace. A simple solution to reduce error would be capping the pace that we predict at $10 m/s$ (this pace represents one of the fastest human sprints, and is highly unlikely to be exceeded by recreational runners).

\begin{figure}[H]
\begin{center}
\includegraphics[width=6cm]{all.png}
\caption{A graph comparing actual pace with predicted pace. }
\end{center}
\end{figure}

However, we have two implicit truths about these predictions. The first is that predicting pace for shorter runs should be easier (more accurate) than doing so for longer runs. This is because pace can vary less over a shorter run, so we would assume less possible variance in our model. The second is that predicting pace in the second half of a run should be more accurate than doing so for the first half of the run. This is because the second half of the run is by default closer to the end, so runners? parameters have less time to change. We decided to test these two assumptions, with the specific goal of making sure that we still have reasonable accuracy over the harder of these two problems. This is also a useful benchmark, as future optimizations should not just attempt to reduce RMSE, but also attempt to increase accuracy for these harder subsets of the problem.\\

In the first case, we compared the accuracy of predicted pace between short runs and long runs. To do so, we repeated the algorithm presented above over a group of runs less than 3km in length as well as over a group of runs greater than 5km in length. We found a RMSE of 0.8771  for the group of short runs as compared to an RMSE of 4.0195 for the group of longer runs.

\begin{figure}[H]
\begin{center}
\includegraphics[width=6cm]{longshort.png}
\caption{A graph comparing actual pace with predicted pace with long runs shown in red and short runs in blue. }
\end{center}
\end{figure}

In the second case, we compared the accuracy of predicted pace between the predictions made in the first half and those made in the second half of runs. We used the same procedure as above, but drew random starting points from the first and second half of time values, respectively. We found a RMSE of 0.5969 for the first halves of runs as compared to an RMSE of 2.9751 for the second halves runs.

\begin{figure}[H]
\begin{center}
\includegraphics[width=6cm]{halves.png}
\caption{A graph comparing actual pace with predicted pace with predictions made in the first half of a run shown in blue and those in the second half of a run in red. }
\end{center}
\end{figure}

@@

\indent This project used a K-nearest-neighbor search attempting to take past runs and combine them in reveal information about how a typical run from a user would go. This estimation is then useful in predicting a good recommendation for how the runner should progress from that moment in the run. This approach to this issue resulted in a good prediction engine for short runs and a good starting point for longer runs. The longer runs are exceedingly difficult to extrapolate to because of the many factors that can come into play over the remaining time limit as well as the limited amount of long distance run data available during this project. Having access to a good amount of RunKeeper's or MapMyRun's data would allow for much more analysis. Combining run results is a difficult task and has many points of improvement that could probably apply to non-running type exercises as well. \\

	This investigation provided useful results, but it was intended mainly as a starting point. The problem of workout recommendation is the source of entire professions (i.e. personal trainers). This project assumed all runs were 'Äúgood'Äù runs and thus should be taken into account for prediction. In reality, some runs are ruined by a bad cramp, an error in data collection, or a phone call mid-run and thus should not be used to describe typical performance. Additionally, we ran our K-nearest-neighbor algorithm on the current run and matched other runs together based on just the Z variable of total run length. This could be improved if we tried to find runs similar to the current one based on the total length as well as the performance so far. This would allow us to, at least to some degree, more properly predict the longer runs as well as fix another major development that this problem needs before anyone can call it solved: atrophy during extended periods of inactivity. This app currently assumes the user is constantly training, however, a very common use case is ignored. Namely, a runner will often take a few weeks or even months off because they are particularly busy or it'Äôs winter. This would lead to a mismatch where the runner is in much worse shape than our current algorithm understands and then tells the runner to essentially sprint the rest of the run. Matching by performance so-far in the run would allow the algorithm to implicitly match on conditioning level as well. With a large enough user base, this could become very powerful as thousands of runs can be combined from a multitude of different users, much in the same way Netflix powers their suggestion engine. Furthermore, these approaches could easily be applied to many more types of exercise such as swimming or weight lifting as well. \\

	With a little bit of improvement, this type of project would be very valuable to an application such as RunKeeper. Taking the expected performance and translating it into a pace that the runner should keep would be a good next step in this project. Using assumptions based on the height of the runner would be a simple way to suggest a running pace based on speed, however an entirely separate study should be done on how to use machine learning to match a specific user to the proper pace based on where they are in the run and any other factors specific to their person and their run. This could then be a powerful tool to give runners the coach they don'Äôt even know they have.
